Differential abundance analysis of taxonomically-biased microbiome measurements

### 2022-02-15 case-studies.Rmd
Make clarifying edits to case studies

@leopold2020host performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of 'priority effects'---the advantage gain by a species from being allowed to colonize first.
Both analyses were based on fold changes in species proportions and so in principle were sensitive to taxonomic bias.
To ensure the results were accurate, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.

Calibration had negligible impact on the results (personal communication with Devin Leopold and confirmed by our own reanalysis).
To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \@ref(fig:leopold2020host-variation)).
Despite the 13X variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62X and a geometric standard deviation of 1.05X.
This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \@ref(fig:leopold2020host-variation)A).
But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3X and by just 1.5X between the two most dominant.
The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual taxa, it had a negligible impact on the inferred fold changes and the DA analyses based on them.

@@

We sought to further understand the implications of the sparsity of gut microbiome for the effect of bias on DA analyses in the context of a real DA analysis.
@vieirasilva2019quan analyzed variation in absolute abundance of genera in stool samples from patients with primary sclerosing cholangitis and/or inflammatory bowel disease.
Absolute abundances of genera were obtained via the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)) with proportions measured from 16S sequencing and total abundance measured from flow cytometry.
The authors the rank-based Spearman correlation to quantify the associations in absolute abundance and fecal calprotectin concentration, a biomarker of intestinal inflammation.
[This will be described in Section \@ref(differential-abundance): Spearman correlation is equivalent to doing a regression on the rank-transformed data and is another form of DA analysis; changes in ranks of proportions are affected by bias, and changes in ranks of ratios are not. So rank-based methods (Wilcoxon for boolean covariate and Spearman for continuous covariate) have the same issue as LFC methods when it comes to analyzing proportions instead of ratios.]
As described in Section \@ref(differential-abundance), taxonomic bias in the MGS measurement has a similar effect on analyses of absolute abundances from the total-abundance normalization as it does no analyses of the raw proportions.
Thus we can consider whether bias might affect the Spearman correlations in the study.
When we re-analyzed the Spearman correlations, we found that they were often driven by presence-absence of the given genus, rather than continuous variation in abundance, as evidenced by the fact that for many species, replacing the absolute genus abundances with 0 or 1 indicating presence or absence often did not affect the Spearman correlation.
For these taxa, we can expect bias to have much less impact on the reported correlation than for taxa for which the correlation was driven by continuous variation.
The reason is that systematic variation of the mean efficiency with the covariate (which was, in contrast, log-normally distributed) will tend to systematically shift the ranks of observations in a taxon with a continuously varying abundance, but not one where the rank correlation is driven by the proportion of non-zero observations.
We confirmed this expectation by investigating the sensitivity of Spearman correlations for the top associations that were driven either by presence/absence variation or continuous variation, and found that the correlation was much more sensitive to bias in the second case.

@@

## Summary and conclusions

>>>>

@leopold2020host performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of 'priority effects'---the advantage gained by a species from being allowed to colonize first.
Both analyses were based on fold changes in species proportions and so in principle were sensitive to taxonomic bias.
To improve accuracy, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.

We repeated the two DA analyses of @leopold2020host with and without calibration and found that the results did not meaningfully differ.
To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \@ref(fig:leopold2020host-variation)).
Despite the 13X variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62X and a geometric standard deviation of 1.05X.
This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \@ref(fig:leopold2020host-variation)A).
But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3X and by just 1.5X between the two most dominant.
The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual taxa, it had a negligible impact on the inferred fold changes and the DA analyses based on them.

@@

We sought to further understand the implications of the sparsity of gut microbiome for the effect of bias on DA analyses in the context of a real DA analysis.^[Some of the results in this paragraph seem to not hold up to more careful investigation. More generally, this paragraph is quite speculative and may get cut from subsequent versions.]
@vieirasilva2019quan analyzed variation in absolute abundance of genera in stool samples from patients with primary sclerosing cholangitis and/or inflammatory bowel disease.
Absolute abundances of genera were obtained via the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)) with proportions measured from 16S sequencing and total abundance measured from flow cytometry.
The authors the rank-based Spearman correlation to quantify the associations in absolute abundance and fecal calprotectin concentration, a biomarker of intestinal inflammation.
[This will be described in Section \@ref(differential-abundance): Spearman correlation is equivalent to doing a regression on the rank-transformed data and is another form of DA analysis; changes in ranks of proportions are affected by bias, and changes in ranks of ratios are not. So rank-based methods (Wilcoxon for boolean covariate and Spearman for continuous covariate) have the same issue as LFC methods when it comes to analyzing proportions instead of ratios.]
As described in Section \@ref(differential-abundance), taxonomic bias in the MGS measurement has a similar effect on analyses of absolute abundances from the total-abundance normalization as it does no analyses of the raw proportions.
Thus we can consider whether bias might affect the Spearman correlations in the study.
When we re-analyzed the Spearman correlations, we found that they were often driven by presence-absence of the given genus, rather than continuous variation in abundance, as evidenced by the fact that for many species, replacing the absolute genus abundances with 0 or 1 indicating presence or absence often did not affect the Spearman correlation.
For these taxa, we can expect bias to have much less impact on the reported correlation than for taxa for which the correlation was driven by continuous variation.
The reason is that systematic variation of the mean efficiency with the covariate (which was, in contrast, log-normally distributed) will tend to systematically shift the ranks of observations in a taxon with a continuously varying abundance, but not one where the rank correlation is driven by the proportion of non-zero observations.
We confirmed this expectation by investigating the sensitivity of Spearman correlations for the top associations that were driven either by presence/absence variation or continuous variation, and found that the correlation was much more sensitive to bias in the second case.

@@

## Summary
